Paradigms in Physics 2.0

### 2017-06-27 paper.tex
add paragraph from Liz

To physicists, mathematics is a language for telling stories about
conceptual physical relationships.  Our understandings of conceptual
connections guide us when we manipulate
equations~\cite{gire2008resources, bing2007cognitive,
  wittmann2015mathematical}[75].  For example, when solving ordinary
differential equations by separating variables, we choose limits of
integration so that the lower (and upper) limits for all variables
describe the same physical state.  We monitor algebraic manipulations
using conceptual understandings of physics quantities. For example, we
make sure that each term in a sum has the same physical dimensions
(e.g. you can't add a frequency to a frequency squared). We make sure
that expressions on both sides of an equals sign have the same
dimensions~\cite{lenz2016dimensional}. We consider limiting cases to
make sure that a derived expression reasonably describes the system in
extreme states~\cite{singh2002physical}. Our geometric understandings
guide the formulation of equations~\cite{gire2014arrows}. Sometimes,
we do not have adequate mathematics to describe complex physical
systems. We use conceptual understandings to make choices about
reasonable idealizations in order to create useful mathematical
models~\cite{bing2007cognitive}.

Being fluent in this physics language involves both (1) translating a
conceptual story into a mathematical representation and (2) discerning
a conceptual story from a mathematical
representation~\cite{tuminaro2007elements}. At the introductory level,
students often see concepts and mathematics as being separate. These
students do not expect equations to help their conceptual
understanding~\cite{adams2006new, redish1998expectations} and have
trouble formulating equations from a conceptual analysis. Students who
do not have mathematical fluency for physics struggle in advanced
physics courses~\cite{manogue2009cognitive}.

We propose a course that explicitly focuses on physics majors'
mathematical fluency and sense-making skills. Introductory courses
begin to develop these aspects, but the primary goals of those courses
tend to be for students to apply simple concepts (Newton's Laws,
energy conservation, etc.) to relatively straightforward and very
idealized problems.

This course is taken either just after or concurrently with the last
quarter of the introductory physics sequence. It leverages students'
understandings of simple mechanical systems from introductory physics
to develop their mathematical fluency in the context of less
idealized, more sophisticated systems. Each physics topic
(velocity-dependent forces, Lagrangian mechanics, Hamiltonian
mechanics, special relativity) will focus on specific math methods
(e.g. differential equations, calculus of variations, generalized
coordinates, Legendre transformations, linear transformations,
hyperbola geometry) and will explicitly feature strategies to make
physical sense of mathematics (symbolic problem solving, reading
equations like sentences, making idealization, coordinating
representations, limiting case analysis, dimensional checking).

>>>>

While the Challenges courses takes a more experimental/applied physics
focus, the Methods of Theoretical Mechanics course has a more
theoretical physics flavor. The theme of this course is the discussion
of strategies for making sense of physics problems and symbolic
problem solving. This sense-making includes coordinating and
interpreting symbolic expressions with conceptual understandings,
geometric relationships, and physical intuitions.

This Theoretical Mechanics course is aimed at students who are taking
or have completed the last course in the introductory sequence. The
physics content of the course is advanced Newtonian mechanics,
introduction to Lagrangian and Hamiltonian techniques, and special
relativity. These topics at this stage for the students are convenient
for explicit discussion of sense-making strategies in physics because
(a) students are have recently studied problems that serve as limiting
cases for more complex problems, (b) students at this level are
transitioning from solving problems primarily involving numbers to
problems with only symbolic parameters, and (c) we can discuss
sense-making as strategies for approaching a problem, evaluating
answers, and refining intuitions about relativistic and other
unfamiliar situations. For example, we teach students how to use
spacetime diagrams to develop a story about a relativistic situation
and how to use hyperbola trig on these diagrams to perform Lorentz
transformations that can be checked against the results of algebraic
Lorentz transformations.  The sense-making goals of the course are on
equal footing with the physics goals, and sense-making is integrated
explicitly in all aspects of the course: exams, homework and during
in-class activities. Our aim is that students will develop
sense-making skills that will improve their learning in advanced
courses and will impress their upper-division course instructors,
research advisors, and future employers.

%% To physicists, mathematics is a language for telling stories about
%% conceptual physical relationships.  Our understandings of conceptual
%% connections guide us when we manipulate
%% equations~\cite{gire2008resources, bing2007cognitive,
%%   wittmann2015mathematical}[75].  For example, when solving ordinary
%% differential equations by separating variables, we choose limits of
%% integration so that the lower (and upper) limits for all variables
%% describe the same physical state.  We monitor algebraic manipulations
%% using conceptual understandings of physics quantities. For example, we
%% make sure that each term in a sum has the same physical dimensions
%% (e.g. you can't add a frequency to a frequency squared). We make sure
%% that expressions on both sides of an equals sign have the same
%% dimensions~\cite{lenz2016dimensional}. We consider limiting cases to
%% make sure that a derived expression reasonably describes the system in
%% extreme states~\cite{singh2002physical}. Our geometric understandings
%% guide the formulation of equations~\cite{gire2014arrows}. Sometimes,
%% we do not have adequate mathematics to describe complex physical
%% systems. We use conceptual understandings to make choices about
%% reasonable idealizations in order to create useful mathematical
%% models~\cite{bing2007cognitive}.

%% Being fluent in this physics language involves both (1) translating a
%% conceptual story into a mathematical representation and (2) discerning
%% a conceptual story from a mathematical
%% representation~\cite{tuminaro2007elements}. At the introductory level,
%% students often see concepts and mathematics as being separate. These
%% students do not expect equations to help their conceptual
%% understanding~\cite{adams2006new, redish1998expectations} and have
%% trouble formulating equations from a conceptual analysis. Students who
%% do not have mathematical fluency for physics struggle in advanced
%% physics courses~\cite{manogue2009cognitive}.

%% We propose a course that explicitly focuses on physics majors'
%% mathematical fluency and sense-making skills. Introductory courses
%% begin to develop these aspects, but the primary goals of those courses
%% tend to be for students to apply simple concepts (Newton's Laws,
%% energy conservation, etc.) to relatively straightforward and very
%% idealized problems.

%% This course is taken either just after or concurrently with the last
%% quarter of the introductory physics sequence. It leverages students'
%% understandings of simple mechanical systems from introductory physics
%% to develop their mathematical fluency in the context of less
%% idealized, more sophisticated systems. Each physics topic
%% (velocity-dependent forces, Lagrangian mechanics, Hamiltonian
%% mechanics, special relativity) will focus on specific math methods
%% (e.g. differential equations, calculus of variations, generalized
%% coordinates, Legendre transformations, linear transformations,
%% hyperbola geometry) and will explicitly feature strategies to make
%% physical sense of mathematics (symbolic problem solving, reading
%% equations like sentences, making idealization, coordinating
%% representations, limiting case analysis, dimensional checking).
